---
title: "ST462_ST662"
subtitle: "Assignment 1"
output:
  html_document: default
  pdf_document:
    number_sections: yes

```{r load_packages1, message = FALSE}
if (!require(gridExtra)) install.packages('gridExtra', dependencies = TRUE) 
library(gridExtra)
if (!require(knitr)) install.packages('knitr', dependencies = TRUE) 
library(knitr)
if (!require(GGally)) install.packages('GGally', dependencies = TRUE)
library(GGally)
if (!require(kableExtra)) install.packages('kableExtra', dependencies = TRUE) 
library(kableExtra)
if (!require(jtools)) install.packages('jtools', dependencies = TRUE) 
library(jtools)
if (!require(rsample)) install.packages('rsample', dependencies = TRUE)
library(rsample)
if (!require(broom)) install.packages('broom', dependencies = TRUE) 
library(broom)
if (!require(tidyverse)) install.packages('tidyverse', dependencies = TRUE)
library(tidyverse)
```
```{r include=FALSE}
if(knitr::is_html_output()){options(knitr.table.format = "html")} else {options(knitr.table.format = "latex")}
```
Adding the file for Q1
Installing package
```{r}
install.packages("ISLR2")
library(ISLR2)

```
Loading the Data 
```{r}
data(Auto)
```

Scatter Matrix plot
```{r}
pairs(Auto)
```

Installing packages for data analysis
```{r, include=FALSE, echo=FALSE}
rm(list=ls()) #clear environment

library(ISLR)
library(DAAG)
library(PerformanceAnalytics)
install.packages("ggplot2")
install.packages("GGally")
install.packages("psych")

#read dataset
auto <- Auto
```

Question A
Scatter plot matrix includes all variables in the "Auto" data set
```{r}
pairs(auto)
```


```{r}
library(ggplot2) 
library(GGally)
Scatter_Matrix <- ggpairs(auto,columns = c(1,6:8), 
                          title = "Scatter Plot Matrix for Auto Dataset", 
                          axisLabels = "show") 
ggsave("Scatter plot matrix.png", Scatter_Matrix, width = 7, 
       height = 7, units = "in") 
Scatter_Matrix


```


```{r, include=TRUE, echo=TRUE}
#exploratory analysis
chart.Correlation(Auto[, -9])
```


Question D
USe the LM function for 
Regression model orignial
```{r}
model <- lm(mpg ~ . - name, data = Auto)
# Print the summary of the regression model
summary(model)
plot(model)
```

Question D
Using the plot function for linear fit

```{r}
par(mfrow = c(2,2))
plot(model)
```

Quesiton E
Adding different transformations (* and :) to models
```{r}
#Squaring

model_square_accel = lm(mpg ~ horsepower + weight + cylinders + origin + acceleration*acceleration, data = auto)
summary(model_square_accel)


#interaction terms
#Horse and weight interaction
model_horseweight <- lm(mpg ~ horsepower + weight + horsepower:weight, data = auto)

summary(model_horseweight)
plot(model_horseweight)

#adding interaction between weight and accel, horse and accel, year and origin

model_triple = lm(mpg ~.-name-cylinders-acceleration+year:origin+displacement:weight+
                  displacement:weight+acceleration:horsepower+acceleration:weight, data=auto)
summary(model_triple)
plot(model_triple)


```


Question F
Changing the models to log-log, sqrt, ^2 and other interaction terms included

```{r}
# Fit a linear regression model with original variables
model_original <- lm(mpg ~ horsepower + weight + cylinders, data = auto)
summary(model_original)

# Fit a linear regression model with log-transformed variables
model_log <- lm(mpg ~ log(horsepower) + log(weight) + log(cylinders), data = auto)
summary(model_log)


# Fit a linear regression model with square root-transformed variables
model_sqrt <- lm(mpg ~ sqrt(horsepower) + sqrt(weight) + sqrt(cylinders), data = auto)
summary(model_sqrt)

# Fit a linear regression model with squared variables
model_squared <- lm(mpg ~ I(horsepower^2) + I(weight^2) + I(cylinders^2), data = auto)
summary(model_squared)

```


